---
title: "Mongolian Vowel harmony analysis code"
author: "Auromita"
date: "10/6/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
# options(scipen=999) # turn off scientific e notation, show full numbers. To turn on: options(scipen=0)

```


Load packages and data, create dataframes from Lobanov normalized formant values, and make subsets for harmonic, non-harmonic, ATR harmony, rounding harmony.

(note: file paths relative to the Rproject root folder (mongolian_vowel_harmony))

```{r}

library(here) # sets working directory to project root folder (default for .rmd files: folder that contains the file)
library(readxl)
library(tidyverse)
library(lme4)
library(phonR)
library(effectsize)  # to get approximations of effect size from test statistics
library(lmerTest) # to get t and F values from lme model

# read in data
df <- read_excel("formant_data/formants_combined.xlsx") 
str(df)

# create "consonant" column (intervening consonant)
df <- df %>% 
  mutate(temp = word, .after = v2) %>%
  #find the vowels in the word, replace with underscores
  mutate(temp = str_replace_all(temp, v1, "_")) %>%
  mutate(temp = str_replace_all(temp, v2, "_")) %>%
  #create column with whatever is between the underscores (intervening consonant)
  separate(temp, into = c(NA,"consonant",NA), sep = "_") %>%
  mutate_if(is.character,as.factor) %>%
  mutate(iteration = as.factor(iteration))

str(df)

attach(df)
# separate objects for F1 and F2 values 

F1 <- cbind(f1_v1_t1,	f1_v1_t2,	f1_v1_t3,	f1_v1_t4,	f1_v1_t5,	f1_v1_t6,	f1_v1_t7,	f1_v1_t8,	f1_v1_t9,	f1_v1_t10, f1_v2_t1,	f1_v2_t2,	f1_v2_t3,	f1_v2_t4,	f1_v2_t5,	f1_v2_t6,	f1_v2_t7,	f1_v2_t8,	f1_v2_t9,	f1_v2_t10)
F2 <- cbind(f2_v1_t1,	f2_v1_t2,	f2_v1_t3,	f2_v1_t4,	f2_v1_t5,	f2_v1_t6,	f2_v1_t7,	f2_v1_t8,	f2_v1_t9,	f2_v1_t10, f2_v2_t1,	f2_v2_t2,	f2_v2_t3,	f2_v2_t4,	f2_v2_t5,	f2_v2_t6,	f2_v2_t7,	f2_v2_t8,	f2_v2_t9,	f2_v2_t10)

# new data frame with normalized formant values (formant-intrinsic normalization)

f1lobanov <- with(df,normLobanov(F1,group = speaker))
f2lobanov <- with(df,normLobanov(F2,group = speaker))
df_lobanov <- data.frame(speaker, iteration,word, harmony_type, v1, v2, consonant, f1lobanov, f2lobanov)

rm(F1, F2, f1lobanov, f2lobanov)

detach(df)


# subsets for harmonic and non-harmonic sequences
df_nhar <- filter(df_lobanov, harmony_type == "non-harmonic")
df_har <- filter(df_lobanov, harmony_type != "non-harmonic")

df_ATR <- filter(df_lobanov, harmony_type == "ATR")
df_rounding <- filter(df_lobanov, harmony_type == "rounding")

```

## Figures

```{r}
## plots

xlim=c(2.5,-2.5)
ylim=c(4,-2)

```

### Plots for full dataset:

V1:

```{r}

with(df_lobanov, plotVowels(f1_v1_t5, f2_v1_t5, 
                        xlim=xlim, ylim=ylim,
                        vowel = v1, #group = harmony_type,
                        plot.tokens = F, plot.means = T, 
                        var.col.by = v1, var.sty.by = v1, pch.means = v1, 
                        ellipse.fill = T, ellipse.line = F,
                        poly.line = TRUE, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                        pretty = T))
```

V2:

```{r}
with(df_lobanov, plotVowels(f1_v2_t5, f2_v2_t5, 
                        xlim=xlim, ylim=ylim,
                        vowel = v2, #group = harmony_type,
                        plot.tokens = F, plot.means = T, 
                        var.col.by = v2, var.sty.by = v2, pch.means = v2, 
                        ellipse.fill = T, ellipse.line = F,
                        poly.line = TRUE, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                        pretty = T))
```


### Plots for non-harmonic sequences:

V1:

```{r}

with(df_nhar, plotVowels(f1_v1_t5, f2_v1_t5, 
                         xlim=xlim, ylim=ylim,
                         vowel = v1, #group = speaker, 
                         plot.tokens = F, plot.means = T, 
                         var.col.by = v1, var.sty.by = NULL, pch.means = v1, 
                         ellipse.fill = T, ellipse.line = F,
                         poly.line = T, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                         pretty = T))

```

V2:

```{r}
with(df_nhar, plotVowels(f1_v2_t5, f2_v2_t5, 
                         xlim=xlim, ylim=ylim,
                         vowel = v2, #group = speaker, 
                         plot.tokens = F, plot.means = T, 
                         var.col.by = v2, var.sty.by = NULL, pch.means = v2, 
                         ellipse.fill = T, ellipse.line = F,
                         poly.line = T, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                         pretty = T))

```

```{r}

# figure for labphon18 abstract: diffusion in vowel space for harmonic vs non-harmonic subsets (all vowels combined: vowel=NULL)


with(df_nhar, plotVowels(f1_v1_t5, f2_v1_t5, 
                         xlim=xlim, ylim=ylim,
                         vowel = NULL, #group = speaker, 
                         plot.tokens = F, plot.means = T, 
                         var.col.by = NULL, var.sty.by = NULL, pch.means = "V1",cex.means = 1.5, 
                         ellipse.fill = T, fill.opacity = 0.3, col = "#FF6633", ellipse.line = F,
                         poly.line = T, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                         pretty = T))

par(new=TRUE)

with(df_nhar, plotVowels(f1_v2_t5, f2_v2_t5, 
                         xlim=xlim, ylim=ylim,
                         vowel = NULL, #group = NULL, 
                         plot.tokens = F, plot.means = T, 
                         var.col.by = v2, var.sty.by = NULL, pch.means = "V2", cex.means = 1.5,
                         ellipse.fill = T, fill.opacity = 0.6, col = "grey", ellipse.line = F,
                         poly.line = T, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                         pretty = T))


```

### Plots for harmonic sequences:

V1:

```{r}
with(df_har, plotVowels(f1_v1_t5, f2_v1_t5, 
                        xlim=xlim, ylim=ylim,
                        vowel = v1, #group = speaker,
                        plot.tokens = FALSE, plot.means = TRUE, 
                        var.col.by = v1, var.sty.by = v1, pch.means = v1, 
                        ellipse.fill = T, ellipse.line = F, 
                        poly.line = TRUE, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                        pretty = T))
```

V2:

```{r}
with(df_har, plotVowels(f1_v2_t5, f2_v2_t5, 
                        xlim=xlim, ylim=ylim,
                        vowel = v2, #group = speaker,
                        plot.tokens = F, plot.means = T, 
                        var.col.by = v2, var.sty.by = v2, pch.means = v2, 
                        ellipse.fill = T, ellipse.line = F,
                        poly.line = TRUE, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                        pretty = T))
```


```{r}

# figure for labphon18 abstract: diffusion in vowel space for harmonic vs non-harmonic subsets (all vowels combined: vowel=NULL)

with(df_har, plotVowels(f1_v1_t5, f2_v1_t5, 
                         xlim=xlim, ylim=ylim,
                         vowel = NULL, #group = speaker, 
                         plot.tokens = F, plot.means = T, 
                         var.col.by = NULL, var.sty.by = NULL, pch.means = "V1",cex.means = 1.5, 
                         ellipse.fill = T, fill.opacity = 0.25, col = "#FF6633", ellipse.line = F,
                         poly.line = T, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                         pretty = T))

par(new=TRUE)

with(df_har, plotVowels(f1_v2_t5, f2_v2_t5, 
                         xlim=xlim, ylim=ylim,
                         vowel = NULL, #group = NULL, 
                         plot.tokens = F, plot.means = T, 
                         var.col.by = v2, var.sty.by = NULL, pch.means = "V2", cex.means = 1.5,
                         ellipse.fill = T, fill.opacity = 0.4, col = "grey", ellipse.line = F,
                         poly.line = T, poly.order = c("i", "e", "a", "ɔ", "o", "ʊ", "u"),
                         pretty = T))

```

## LME models

For linear mixed models, the data was mean-centered and scaled prior to analysis, as this increases the likelihood of convergence (Bates et al. 2015a, Eager&Roy 2017):

```{r}

df_lobanov.CS <- df_lobanov %>% mutate_if(is.numeric,scale)
df_har.CS <- df_har %>% mutate_if(is.numeric,scale)
df_nhar.CS <- df_nhar %>% mutate_if(is.numeric,scale)


# rm(df,df_har,df_nhar,df_rounding,df_ATR)

```


We expect that F1 of a vowel is predicted by the vowel identity. We want to test the hypothesis that F1 is also predicted by the identity of the other vowel.

Model names:
- anticipatory model: v1 explained by v2
- carryover model: v2 explained by v1

Below, we compare the full models for anticipatory and carryover coarticulation to their corresponding null models.


## Models for F1

### anticipatory model (effect of v2 on v1)

```{r}

# full model
ant_vh.model = lmer(f1_v1_t5 ~ 
                     v1+ v2 +
                     (1|word), 
                     data = df_lobanov.CS, REML = FALSE)
summary(rePCA(ant_vh.model))
summary(ant_vh.model)

```

```{r}
# null model
ant_null.model = lmer(f1_v1_t5 ~ 
                       v1 + 
                       (1|word), 
                       data = df_lobanov.CS, REML = FALSE)
summary(rePCA(ant_null.model))
summary(ant_null.model)


anova(ant_vh.model, ant_null.model)

```

This shows that v2 reliably predicts v1, showing coarticulation in the right-to-left (anticipatory) direction.


### carryover model (effect of v1 on v2)
```{r}

# full model
car_vh.model = lmer(f1_v2_t5 ~ 
                     v2 + v1 +
                     (1|word), 
                     data = df_lobanov.CS, REML = FALSE)
summary(rePCA(car_vh.model))
summary(car_vh.model)

# null model
car_null.model = lmer(f1_v2_t5 ~
                       v2 +
                       (1|word), 
                       data = df_lobanov.CS, REML = FALSE)
summary(car_null.model)


anova(car_vh.model, car_null.model)

```

This shows that the v1 reliably predicts v2, showing coarticulation in the left-to-right direction.

Comparing the effect size of the explanatory variable (identity of the other vowel) in the anticipatory vs the carryover model:

```{r}

# Anticipatory model: estimating effect size (partial eta squared) for the main variable (v2) from F-value (package: "effectsize")

anova <- anova(ant_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
ant_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)

#---

# Carryover model: estimating effect size (partial eta squared) for the main variable (v1) from F-value 
anova <- anova(car_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
car_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)

# compare effect sizes
ant_greater <- ant_effect > car_effect

if(ant_greater == TRUE) {
   print("Anticipatory effect greater than carryover.")
} else {
   print("Carryover effect is greater than anticipatory.")
}

```
--> The effect is larger in the anticipatory (right-to-left) direction.



## Interaction model:

We want to see if patterns of coarticulation differ across harmonic vs non-harmonic vowel sequences, i.e. whether the predictive power of the other vowel is mediated by harmony type. The next section compares the present full model to one where harmony_type is added as an interaction term.

```{r}

# coding harmony type as a binary variable (harmonic vs non-harmonic) and removing empty levels
df_harmonytype.CS <- df_lobanov.CS %>%
  mutate(harmony_type = str_replace_all(harmony_type, "ATR|rounding", "harmonic")) %>%
  mutate_if(is.character,as.factor)

df_har.CS <- df_har.CS %>%
  mutate(harmony_type = str_replace_all(harmony_type, "ATR|rounding", "harmonic")) %>%
  mutate_if(is.character,as.factor)

df_nhar.CS$harmony_type <- droplevels(df_nhar.CS$harmony_type)


```

### Anticipatory model (effect of v2 on v1)

```{r, eval=FALSE}

# interaction model for anticipatory coarticulation
ant_vh.model = lmer(f1_v1_t5 ~ 
                     v1 + harmony_type*v2 + 
                     (1|word), 
                   data = df_harmonytype.CS, REML = FALSE)
summary(ant_vh.model)

# null (non-interaction) model
ant_null.model = lmer(f1_v1_t5 ~ 
                       v1 + v2 +
                       (1|word), 
                     data = df_harmonytype.CS, REML = FALSE)
summary(ant_null.model)


anova(ant_vh.model, ant_null.model)

```

The harmony type*F1_v2 interaction has a significant effect in the full model, and improves model fit. That is, anticipatory coarticulation is moderated by harmony type.


### Carryover model (effect of v1 on v2)

```{r}

# interaction model for carryover coarticulation
car_vh.model = lmer(f1_v2_t5 ~ 
                     v2 + harmony_type*v1 +
                     (1|word), 
                   data = df_harmonytype.CS, REML = FALSE)
summary(car_vh.model)

# null (non-interaction) model
car_null.model = lmer(f1_v2_t5 ~
                       v2 + v1 +
                       (1|word), 
                     data = df_harmonytype.CS, REML = FALSE)
summary(car_null.model)


anova(car_vh.model, car_null.model)

```
The harmony type*F1_v1 interaction has a significant effect in the full model. That is, carryover coarticulation is moderated by harmony type. 

--> however, adding the interaction term does not significantly improve model fit 

Overall, extent and patterns of coarticulation are different across harmonic vs. non-harmonic sequences. The next section compares coarticulation in harmonic vs. non-harmonic words.

---

## Subsetting data by harmony type


### subset: non-harmonic

### anticipatory model (effect of v2 on v1)

```{r}

# full model

#---> by-word random intercept gives singular fit; rePCA shows that there is by-speaker variation here
ant_vh.model = lmer(f1_v1_t5 ~ 
                     v1+ v2 +
                     (1|speaker), 
                   data = df_nhar.CS, REML = FALSE)
summary(rePCA(ant_vh.model))
summary(ant_vh.model)



# null model
ant_null.model = lmer(f1_v1_t5 ~ 
                       v1 + 
                       (1|speaker), 
                     data = df_nhar.CS, REML = FALSE)
summary(ant_null.model)

#summary(rePCA(ant_null.model))


# comparing full and null models
anova(ant_vh.model, ant_null.model)

```
Anticipatory coarticulation is robust in non-harmonic sequences.


### Carryover model (effect of v1 on v2)

```{r}

# full model
car_vh.model = lmer(f1_v2_t5 ~ 
                     v2 + v1 +
                     (1|speaker), 
                   data = df_nhar.CS, REML = FALSE)
summary(rePCA(car_vh.model))
summary(car_vh.model)

# null model
car_null.model = lmer(f1_v2_t5 ~
                       v2 + 
                       (1|speaker), 
                     data = df_nhar.CS, REML = FALSE)
summary(car_null.model)

anova(car_vh.model,car_null.model)

```
No significant difference; carryover coarticulation is not robust in non-harmonic sequences.


Comparing effect sizes in the anticipatory vs carryover models:

```{r}
# Anticipatory model: estimating effect size (partial eta squared) for the main variable (F1 of v2) from F-value (package: "effectsize")

anova <- anova(ant_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
ant_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)
#---

# Carryover model: estimating effect size (partial eta squared) for the main variable (F1 of v1) from F-value 
anova <- anova(car_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
car_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)

# compare effect sizes
ant_greater <- ant_effect > car_effect

if(ant_greater == TRUE) {
   print("Anticipatory effect greater than carryover.")
} else {
   print("Carryover effect is greater than anticipatory.")
}


```
In non-harmonic sequences, the effect size of anticipatory coarticulation (opposite todirection of vowel harmony) is greater than the carryover coarticulation.

--> difference between anticipatory and carryover is greater in non-harmonic sequences? - check -- 0.169


### subset: harmonic

### Anticipatory model (effect of v2 on v1)

```{r}

# full model
ant_vh.model = lmer(f1_v1_t5 ~ 
                     v1+ v2 +
                     (1|word), 
                   data = df_har.CS, REML = FALSE)
summary(ant_vh.model)

# null model
ant_null.model = lmer(f1_v1_t5 ~ 
                       v1 +
                       (1|word), 
                     data = df_har.CS, REML = FALSE)
summary(ant_null.model)


anova(ant_vh.model, ant_null.model)

```
The full model is significantly better than the null model; anticipatory coarticulation in robust in harmonic sequences.


### Carryover model (effect of v1 on v2)

```{r}

car_vh.model = lmer(f1_v2_t5 ~ 
                     v2 + v1 +
                     (1|word), 
                     data = df_har.CS, REML = FALSE)
summary(car_vh.model)

car_null.model = lmer(f1_v2_t5 ~
                       v2 +
                       (1|word), 
                       data = df_har.CS, REML = FALSE)
summary(car_null.model)


anova(car_vh.model, car_null.model)

```
The full model is significantly better than the null model; carryover coarticulation in robust in harmonic sequences.


Comparing effect sizes in the anticipatory vs carryover models:

```{r}
# Anticipatory model: estimating effect size (partial eta squared) for the main variable (F1 of v2) from F-value (package: "effectsize")

anova <- anova(ant_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
ant_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)
#---

# Carryover model: estimating effect size (partial eta squared) for the main variable (F1 of v1) from F-value 
anova <- anova(car_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
car_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)

# compare effect sizes
ant_greater <- ant_effect > car_effect

if(ant_greater == TRUE) {
   print("Anticipatory effect greater than carryover.")
} else {
   print("Carryover effect is greater than anticipatory.")
}


```
--> in harmonic sequences-- effect sizes for coarticulation in both directions is greater than in non-harmonic subset

## Models for F2

### Anticipatory model (effect of v2 on v1)

```{r}

# full model
ant_vh.model = lmer(f2_v1_t5 ~ 
                     v1+ v2 +
                    (1|word), 
                     data = df_lobanov.CS, REML = FALSE)
summary(rePCA(ant_vh.model))
summary(ant_vh.model)

# null model
ant_null.model = lmer(f2_v1_t5 ~ 
                       v1 +
                       (1|word), 
                       data = df_lobanov.CS, REML = FALSE)
summary(ant_null.model)


anova(ant_vh.model, ant_null.model)

```

The full model is marginally better than the null model, showing coarticulation in the right-to-left (anticipatory) direction.

### Carryover model (effect of v1 on v2)
```{r}

# full model
car_vh.model = lmer(f2_v2_t5 ~ 
                     v2 + v1 +
                     (1|word), 
                     data = df_lobanov.CS, REML = FALSE)
summary(car_vh.model)

# null model
car_null.model = lmer(f2_v2_t5 ~
                       v2 +
                       (1|word), 
                       data = df_lobanov.CS, REML = FALSE)
summary(car_null.model)


anova(car_vh.model, car_null.model)

```

This shows that the v1 reliably predicts the F2 of v2, showing coarticulation in the left-to-right direction.

Comparing effect sizes in the anticipatory vs carryover models:

```{r}
# Anticipatory model: estimating effect size (partial eta squared) for the main variable (F1 of v2) from F-value (package: "effectsize")

anova <- anova(ant_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
ant_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)
#---

# Carryover model: estimating effect size (partial eta squared) for the main variable (F1 of v1) from F-value 
anova <- anova(car_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
car_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)

# compare effect sizes
ant_greater <- ant_effect > car_effect

if(ant_greater == TRUE) {
   print("Anticipatory effect greater than carryover.")
} else {
   print("Carryover effect is greater than anticipatory.")
}

```
Effect size is greater in the left-to-right direction; same as that of vowel harmony.


## Interaction model:

We want to see if the patterns of coarticulation differ across harmonic vs non-harmonic vowel sequences.The next section compares the present full model to one where harmony_type is added as an interaction term.

### Anticipatory model (effect of v2 on v1)

```{r}

# interaction model
ant_vh.model = lmer(f2_v1_t5 ~ 
                     v1 + harmony_type*v2 +
                     (1|word), 
                   data = df_harmonytype.CS, REML = FALSE)
summary(ant_vh.model)

# null (non-interaction) model
ant_null.model = lmer(f2_v1_t5 ~ 
                       v1 + v2 +
                       (1|word), 
                     data = df_harmonytype.CS, REML = FALSE)
summary(ant_null.model)


anova(ant_vh.model, ant_null.model)

```

Adding harmony type as an interaction term improves model fit; extent of anticipatory coarticulation in F2 differs across harmonic vs non-harmonic sequences.


### Carryover model (effect of v1 on v2)

```{r}

# interaction model
car_vh.model = lmer(f2_v2_t5 ~ 
                     v2 + harmony_type*v1+
                     (1|word), 
                   data = df_harmonytype.CS, REML = FALSE)
summary(car_vh.model)

# null (non-interaction) model
car_null.model = lmer(f2_v2_t5 ~
                       v2 + v1 +
                       (1|word), 
                     data = df_harmonytype.CS, REML = FALSE)
summary(car_null.model)


anova(car_vh.model, car_null.model)

```

Adding harmony type as an interaction term does not improve model fit; extent of carryover coarticulation in F2 does not differ across harmonic vs non-harmonic sequences.

Overall, there are differences in coarticulatory patterns between harmonic and non-harmonic sequences. To examine this difference, looking at coarticulation in harmonic and non-harmonic subsets:


## Subsetting data by harmony type


### subset: non-harmonic

### Anticipatory model (effect of v2 on v1)

```{r}

# full model

ant_vh.model = lmer(f2_v1_t5 ~ 
                     v1+ v2 + 
                     (1|speaker), 
                   data = df_nhar.CS, REML = FALSE)
summary(rePCA(ant_vh.model))
summary(ant_vh.model)

# null model
ant_null.model = lmer(f2_v1_t5 ~ 
                       v1 + 
                       (1|speaker) + (1|word), 
                     data = df_nhar.CS, REML = FALSE, control = lmerControl(optimizer = "bobyqa"))
summary(ant_null.model)
summary(rePCA(ant_null.model))

anova(ant_vh.model, ant_null.model)

```
V2 terms in the full model are significant, suggesting coarticulation in the anticipatory direction. However, we cannot compare this to the corresponding null model using an anova, because of convergence issues-- simplifying the random effects structure until models converge give different random effects structures for the full and null model.


### Carryover model (effect of v1 on v2)

```{r}

# full model
car_vh.model = lmer(f2_v2_t5 ~ 
                     v2 + v1 + 
                     (1|speaker), 
                   data = df_nhar.CS, REML = FALSE)
summary(car_vh.model)


# null model
car_null.model = lmer(f2_v2_t5 ~ 
                     v2 +  
                     (1|speaker), 
                   data = df_nhar.CS, REML = FALSE)

summary(car_null.model)

anova(car_vh.model,car_null.model)

```
Full model is not significantly better; carryover coarticulation in F2 is not robust in non-harmonic sequences.


Comparing effect sizes in the anticipatory vs carryover models:

```{r}
# Anticipatory model: estimating effect size (partial eta squared) for the main variable (F1 of v2) from F-value (package: "effectsize")

anova <- anova(ant_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
ant_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)
#---

# Carryover model: estimating effect size (partial eta squared) for the main variable (F1 of v1) from F-value 
anova <- anova(car_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
car_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)

# compare effect sizes
ant_greater <- ant_effect > car_effect

if(ant_greater == TRUE) {
   print("Anticipatory effect greater than carryover.")
} else {
   print("Carryover effect is greater than anticipatory.")
}

```

Effect size of anticipatory coarticulation is greater than carryover coarticulation in non-harmonic sequences.

---------------------------


### subset: harmonic


### Anticipatory model (effect of v2 on v1)

```{r}

# full model
ant_vh.model = lmer(f2_v1_t5 ~ 
                     v1+ v2 + 
                     (1|word), 
                   data = df_har.CS, REML = FALSE)
summary(ant_vh.model)

# null model
ant_null.model = lmer(f2_v1_t5 ~ 
                       v1 + 
                       (1|word), 
                     data = df_har.CS, REML = FALSE)
summary(ant_null.model)


anova(ant_vh.model, ant_null.model)

```
The full model is not significantly better; harmonic sequences do not show anticipatory coarticulation in F2.


### Carryover model (effect of v1 on v2)

```{r}

# full model
car_vh.model = lmer(f2_v2_t5 ~ 
                     v2 + v1 + 
                     (1|word), 
                     data = df_har.CS, REML = FALSE)
summary(car_vh.model)

# null model
car_null.model = lmer(f2_v2_t5 ~
                       v2 +
                       (1|word), 
                       data = df_har.CS, REML = FALSE)
summary(car_null.model)


anova(car_vh.model, car_null.model)

```
The full model is significantly better; carryover coarticulation in F2 is robust in harmonic sequences.


Comparing effect sizes in the anticipatory vs carryover models:

```{r}
# Anticipatory model: estimating effect size (partial eta squared) for the main variable (F1 of v2) from F-value (package: "effectsize")

anova <- anova(ant_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
ant_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)
#---

# Carryover model: estimating effect size (partial eta squared) for the main variable (F1 of v1) from F-value 
anova <- anova(car_vh.model)
F <- anova$`F value`[[2]] %>% signif(digits=3)
df <- anova$`NumDF`[[2]] %>% signif(digits=3)
dendf <- anova$`DenDF`[[2]] %>% signif(digits=3)

# partial eta squared value for explanatory variable: F_to_eta2(Fvalue, NumDF, DenDF)
car_effect <- F_to_eta2(F,df,dendf)$Eta2_partial %>% signif(digits=3) 

rm(anova,F,df,dendf)

# compare effect sizes
ant_greater <- ant_effect > car_effect

if(ant_greater == TRUE) {
   print("Anticipatory effect greater than carryover.")
} else {
   print("Carryover effect is greater than anticipatory.")
}

```

Carryover coarticulation (same direction as vowel harmony) is greater than anticipatory coarticulation for F2 in harmonic sequences.



For F2, could not compare full vs null models for anticipatory coarticulation in non-harmonic sequences. Apart from that, patterns are identical to F1.


